---
title: "Reproducible Research - Week 2"
author: "Vladimir Orlyanchik"
date: "July 26, 2017"
output: html_document
---

reading in the dataset
```{r reading, echo=TRUE}
library(ggplot2)
library(gridExtra)
Data<-read.csv("activity.csv")
```

Processing the data to be suitable for plotting and analysis - removing rowes with NA

```{r processing, echo=TRUE }
TotalStepsvsDate<-aggregate(steps~date, data=Data[complete.cases(Data),], sum)
```

Histogram of the total number of steps taken each day

```{r histogram1, echo=TRUE,cache = TRUE}
hist(TotalStepsvsDate$steps, breaks=15,xlab ="Total number of Steps")
```

Calculations of mean and median ignoring rows with NA

```{r calculations, echo=TRUE}
mean(TotalStepsvsDate$steps)

median(TotalStepsvsDate$steps)

```

Preparing data showing number of steps averaged accross all days for each of 5-min intervals ignoring rows with NA

```{r steps vs interval, echo=TRUE}
TotalStepsvsInterval<-aggregate(steps~interval, data=Data[complete.cases(Data),], mean)
```

The time interval containing maximum average number of steps is:

```{r maximum, echo=TRUE}
which.max(TotalStepsvsInterval$steps)
```

Here is the plot of the Average Number of Steps as a function of 5-min time interval:

```{r time series, echo=TRUE,cache = TRUE}
plot(TotalStepsvsInterval$steps, type="l",xlab = "5-min Interval",ylab = "Average Number of Steps")
```


Calculation of total number of missing (NA) values:

```{r missing values, echo=TRUE}
dim(Data)[[1]]-dim(Data[complete.cases(Data),])[[1]]
```

There are 2304 missing values. That is about 13% of total Data. Since each days contains 288 5-min intervals, there are total 8 days of missing data. One of the simplest ways to substitute the missing data is by using mean value of corresponding 5-min interval calculated ommiting rowes with missing values.

```{r missing value substitution, echo=TRUE}
NewData<-merge(Data,TotalStepsvsInterval,by = "interval",all = TRUE )
bad<-is.na(NewData$steps.x)
NewData$steps.x[bad]<-NewData$steps.y[bad]
TotalStepsvsDateFull<-aggregate(steps.x~date, data=NewData, sum)
```

The values of mean and median

```{r new mean and median, echo=TRUE}
mean(TotalStepsvsDateFull$steps.x)
median(TotalStepsvsDateFull$steps.x)
```
of the modified data set are very close to what we calculated before

The histogram for the new new data set looks very similar as well

```{r histogram2, echo=TRUE,cache = TRUE}
require(gridExtra)
plot1<-qplot(TotalStepsvsDate$steps, main ="W/O missing data", xlab ="Total number of Steps")
plot2<-qplot(TotalStepsvsDateFull$steps.x, main = "Missing data are imputed",xlab ="Total number of Steps")
grid.arrange(plot1, plot2, nrow=2)
```

Calculation of difference in activity pattern between weekdays and weekends

```{r weekday-weekend data prep , echo=TRUE}
NewData$Weekday<-as.factor(ifelse(weekdays(as.Date(NewData$date)) %in% c("Monday", "Tuesday", "Wednesday", "Thursday", "Friday"), "weekday", "weekend"))
day_activity<-aggregate(steps.x~interval+Weekday,data=NewData,mean)
```

Here is the plot of the Average Number of Steps as a function of 5-min time interval for weekdays and weekends:

```{r weekday-weekend plot, echo=TRUE,cache = TRUE}
qplot((seq_along(day_activity$steps.x)) %% 288,day_activity$steps.x,data=day_activity,geom="line", facets=Weekday~., xlab = "5-min Interval", ylab = "Average Number of Steps")
```

During the weekend an active period starts a little bit later in the day. More 
